% !TeX program = lualatex

% setup
\documentclass[
hyperref={bookmarks=false},
xcolor={dvipsnames,svgnames*,x11names*}, 
12pt
]{beamer}
\usepackage{../beamertheme/beamerinnerthemetb}
\usepackage{../beamertheme/beamerouterthemetb}
\usepackage{../beamertheme/beamercolorthemetb}
\usepackage{../beamertheme/beamerthemetb}

% packages
\usepackage{emoji}
\usepackage{graphicx}
\usepackage{xurl}
\usepackage{epigraph}
\usepackage{listings}
\setlength\epigraphwidth{0.7\linewidth}

\definecolor{Rstring}{HTML}{036A07}
\definecolor{Rcomment}{HTML}{4C886B}
\lstset{
	basicstyle=\color{black}\ttfamily\scriptsize,
	backgroundcolor=\color{white}, 
	breaklines=true, 
	keepspaces=true, 
	escapechar={|},
	breakindent=0pt,
	linewidth=1.05\linewidth
}

% options
\setlength{\leftmargini}{0cm}
\hypersetup{
	pdfauthor={Andrea Gilardi},
	colorlinks=true,
	urlcolor=Blue
}

%% Define new stuff for slide numbers
\setbeamertemplate{navigation symbols}{}

\pgfkeys{/visual counter/.cd,
	thickness/.store in=\thickness,
	thickness=0.4ex,
	radius/.store in=\radius,
	radius=1.5ex,
	segment distance/.store in=\segdist,
	segment distance=8,
	color current frame/.store in=\colcurrframe,
	color current frame=orange,
	color old frame/.store in=\cololdframe,
	color old frame=blue,
	color next frame/.store in=\colnextframe,
	color next frame=gray!30,
	color page number/.store in=\colpagenum,
	color page number=white,
	current value/.store in=\currentv,
	current value=1,
	total value/.store in=\totalv,
	total value=2,
	circled page number/.code={
		\begin{tikzpicture}[fill color/.style={}]
			\pgfkeys{/visual counter/.cd, 
				current value=\insertframenumber,
				total value=\inserttotalframenumber,
			}
			\pgfmathtruncatemacro\current{\currentv+1}
			\def\tot{\totalv}
			\def\radiusout{\radius}
			\def\radiusin{\radius-\thickness}
			
			\foreach \s in {1,...,\tot}
			{
				\ifnum\s>\current%
				\tikzset{fill color/.append style={\colnextframe}}%
				\fi%
				\ifnum\s=\current%
				\tikzset{fill color/.append style={\colcurrframe}}%
				\fi%
				\ifnum\s<\current%
				\tikzset{fill color/.append style={\cololdframe}}%
				\fi%
				\fill[fill color]
				({90-360/\tot * (\s - 1)-\segdist}:\radiusout) arc 
				({90-360/\tot * (\s - 1)-\segdist}:{90-360/\tot * (\s)+\segdist}:\radiusout) --
				({90-360/\tot * (\s)+\segdist}:\radiusin) arc 
				({90-360/\tot * (\s)+\segdist}:{90-360/\tot * (\s - 1)-\segdist}:\radiusin);
				% new addition
				\node[inner sep=0pt,text=\colpagenum] at (0,0){\insertframenumber};
			}
		\end{tikzpicture}
	},
}

\setbeamertemplate{footline}{
	\begin{beamercolorbox}[wd=0.95\textwidth, ht=2ex,dp=1ex,sep=1ex]{footline}
		\hfill%
		\tikz\node[/visual counter/.cd,
		segment distance=-2pt,
		radius=0.33cm, thickness=0.33cm,
		color old frame=black,
		color current frame=black!80!gray!50,
		color next frame=black!80!gray!50,
		circled page number,
		]{};
	\end{beamercolorbox}
}
%% End stuff for slide numbers


% metadata
\title{R4DS - Unit 3: R Packages}
\author{Andrea Gilardi}
\date{\today}

\begin{document}
\inserttitlepage
	
\begin{frame}{Outline and main concepts}
\vspace{-0.5cm}
\begin{itemize}
	\itemsep 3ex
	\item The objective of this unit is to present the basic structure of an R package and a set of simple but powerful routines that can be used to build our first package. 
	\item I will showcase the relevant tools and, at the end, we will build an R package that can be used to decide the optimal date for an happy hour. 
\end{itemize}
\end{frame}	


\begin{frame}{What is an R package?}
	\vspace{-0.5cm}
	\begin{itemize}
		\itemsep 3ex
		\item Following the \href{https://r-pkgs.org/}{\emph{R Package - 2e}} book, we could say that \emph{An \textbf{R package} is the fundamental unit of shareable R code. A package bundles together code, data, documentation, and tests, and is easy to share with others.} 
		\item An R package can be stored on CRAN (Comprehensive R Archive Network), whereas its development version can be stored on Github or other hosting services 
		\item R packages are organised in a standardised format that we must follow. Organising code always makes your life easier since we can follow a template. 
	\end{itemize}
\end{frame}	


\begin{frame}{What is an R package? (cont)}
	\vspace{-0.5cm}
	\begin{itemize}
		\itemsep 2ex
		\item A bit of terminology now\dots\,A \textbf{package} is a directory of files which extend R containing, at minimum, the files DESCRIPTION and NAMESPACE and an R/ directory. 
		\item \textbf{A package is not a library}. 
		\item Beware, maintaining and updating an R package can be an extremely time consuming process\dots 
	\end{itemize}
	\vspace{-0.25cm}
	\epigraph{Maybe I'll become a theoretician. Nobody expects you to maintain a theorem.}{Doug Bates (about keeping both Matrix and RcppEigen in sync with CHOLMOD) lme4-author (September 2013)}
\end{frame}

\begin{frame}[fragile]{Peek at the desired product}
\vspace{-0.5cm}
\begin{itemize}
\itemsep 2ex
\item Now we are going to develop an R package named \texttt{statsAndBooze}. The objective of this package is to find the optimal date for happy hour given a set of constraints. 
\item So, for example, 
\begin{lstlisting}
|\textcolor{blue}{library}|(statsAndBooze)
beer_dates <- parse_dates(
  dates = list(
  andrea = |\textcolor{Rstring}{"2023-03-27 / 2023-04-01"}| |\textcolor{Rcomment}{\# available from 27/3 to 1/4}|
  federico = c(|\textcolor{Rstring}{"2023-03-29"}|, |\textcolor{Rstring}{"2023-04-02"}|) |\textcolor{Rcomment}{\# available on 2 days}|
  )
)
decide_happy_hour(beer_dates)
[1] "2023-03-29"
\end{lstlisting}
\end{itemize}
\end{frame}


\begin{frame}[fragile]{Let's start from scratches...}
\vspace{-0.5cm}
\begin{itemize}
\itemsep 2ex
\item During this class we are going to use the R package \texttt{devtools}, so please take a moment to check if it is already installed and, if necessary, install it. 
\item The following code can be used to generate the skeleton of an empty R package named \texttt{packageName}:
\begin{lstlisting}
|\textcolor{blue}{library}|(devtools)
create_package(|\textcolor{Rstring}{"path/for/the/R/packageName"}|)
\end{lstlisting}
So, for example, I'm going to run
\begin{lstlisting}
|\textcolor{blue}{library}|(devtools)
create_package(|\textcolor{Rstring}{"D:/git/statsAndBooze"}|)
\end{lstlisting}
\item The chosen path should point to a non-existing directory that will be created by Rstudio. Do not store an R package inside another R package or a Git repo. 
\end{itemize}
\end{frame}

\begin{frame}[fragile]{Let's start from scratches\dots\;(cont)}
\vspace{-0.5cm}
\begin{itemize}
\itemsep 1ex
\item The previous command should open a new Rstudio session that contains the skeleton of an empty R package. We will explore its content in a couple of minutes.  
\item You should also see a log message like
\begin{lstlisting}
v Creating 'D:/git/statsAndBooze/'
v Setting active project to 'D:/git/statsAndBooze'
v Creating 'R/'
v Writing 'DESCRIPTION'
Package: statsAndBooze
Title: What the Package Does (One Line, Title Case)
Version: 0.0.0.9000
Authors@R (parsed):
* First Last <first.last@example.com> [aut, cre] (YOUR-ORCID-ID)

[truncated...]
\end{lstlisting}
Now we can analyse the log more precisely together! 
\end{itemize}
\end{frame}

\begin{frame}{Let's start from scratches\dots\, (cont)}
\vspace{-0.5cm}
The following lists the content of the new directory: 
\begin{itemize}
\item \texttt{.gitignore}: The file used to control Git versioning; 
\item \texttt{.Rbuildignore}: Similarly to the \texttt{.gitignore} file, this file can be used to exclude some files from the package BUILD. 
\item \texttt{DESCRIPTION}: Stores the metadata of your package (e.g. author, description, dependencies, \dots)
\item \texttt{R/}: The directory where we put the R scripts. 
\item \texttt{NAMESPACE}: declares the functions your package exports and the external functions your package imports from other packages. \textbf{DO NOT EDIT BY HAND}. 
\end{itemize}
Check \href{https://r-pkgs.org/whole-game.html\#create_package}{here} for more details. 
\end{frame}

\begin{frame}{Edit the \texttt{DESCRIPTION}}
\vspace{-0.5cm}
Now we can edit the \texttt{DESCRIPTION} file:
\begin{itemize}
\itemsep 1ex
\item \texttt{Title:} \texttt{Find the best day to have a beer!};
\item \texttt{Version:} I refer you to \url{https://semver.org/}; 
\item \texttt{Authors@R}: Just add name, surname, email, and ORCID; 
\item \texttt{License}: I refer you to \url{https://choosealicense.com/};
\item \texttt{Description:} We are going to fill it at the end.
\end{itemize}
In case you are developing an R package with some co-authors, you need to list their names (and roles) in the Authors field. See \href{https://r-pkgs.org/description.html\#sec-description-authors-at-r}{here} for more details. 
\end{frame}

\begin{frame}[fragile]{Package dependencies}
\vspace{-0.5cm}
\begin{itemize}
\itemsep 2ex
\item As already mentioned, we are trying to develop an R package that parses dates and intervals. 
\item Working with dates is a \href{https://lubridate.tidyverse.org/articles/lubridate.html\#if-anyone-drove-a-time-machine-they-would-crash}{nightmare}\dots\, Therefore, instead of defining our own routines, we are going to create wrappers to another package: \texttt{lubridate}!
\item Please notice that when you develop an R package, you cannot use \texttt{\textcolor{blue}{library}()} since that works only for interactive scripting. See also \href{https://r-pkgs.org/dependencies-mindset-background.html}{here} and \href{https://r-pkgs.org/dependencies-in-practice.html}{here} for more details.  
\item Instead, we can use the function \texttt{use\_package(\textcolor{Rstring}{"pkg"})}. 
\end{itemize}
\end{frame}

\begin{frame}[fragile]{Package dependencies}
\vspace{-0.5cm}
\begin{itemize}
\itemsep 2ex
\item You should now see the following output: 
\begin{lstlisting}
> use_package(|\textcolor{Rstring}{"lubridate"}|)
v Adding 'lubridate' to Imports field in DESCRIPTION
* Refer to functions with `lubridate::fun()`
\end{lstlisting}
The message highlights that whenever we refer to a lubridate function, we need to add the \texttt{lubridate::} prefix. 
		
\item Check also the DESCRIPTION file and see what happened. 
\item Please notice that the same behaviour must be applied to any function which is not included into the \textbf{base} package. 
\item \textbf{Question:} How can you determine which package defines a function? 
\end{itemize}
\end{frame}

\begin{frame}[fragile]{Interactive development}
\vspace{-0.5cm}
\begin{itemize}
\itemsep 2ex
\item Usually, it is much easier if you run the first tests into an interactive session before writing into the package. 
\item Our first objective is to define a function which takes in input a list of strings and returns the parsed dates:
\begin{lstlisting}
parse_dates(
  dates = list(
  andrea = |\textcolor{Rstring}{"2023-03-29"}|, |\textcolor{Rcomment}{\# exactly with this format}|
  federico = |\textcolor{Rstring}{"2023-03-30"}| |\textcolor{Rcomment}{\# exactly with this format}|
  )
)
$andrea
[1] "2023-03-29" |\textcolor{Rcomment}{\# NB: it must have class = "Date"}|
		
$federico
[1] "2023-03-30"
\end{lstlisting}
\item Now it's your turn! Try to code such a function. 
\end{itemize}
\end{frame}

\begin{frame}[fragile]{The first function}
\vspace{-0.5cm}
\begin{itemize}
\itemsep 2ex
\item Now that we have sketched the skeleton of the function, we can add it to our package. First, we need to create an R script into the R/ folder. We can run \texttt{use\_R(\textcolor{Rstring}{"path.R"})}. 
\item So, for example, I'm going to run 
\begin{lstlisting}
use_R(|\textcolor{Rstring}{"parse.R"}|)
v Setting active project to 'D:/git/statsAndBooze'
* Modify 'R/parse.R'
* Call `use_test()` to create a matching test file
\end{lstlisting}
\item We copy the function definition (\textbf{and only the function definition}) into the new script. Every time that we refer to a lubridate function, we need to add \texttt{lubridate::}. 
\end{itemize}
\end{frame}

\begin{frame}[fragile]{The first function (cont)}
\vspace{-0.5cm}
\begin{itemize}
\itemsep 2ex
\item The \texttt{parse.R} file might look like 
\begin{lstlisting}
1. parse_dates <- |\textcolor{blue}{function}|(x) {
2.   lapply(x, lubridate::as_date)
3. }
4. 
\end{lstlisting}
\item Now we can save the file and restart the R session. Then, if you want to make \texttt{parse\_dates()} available for testing, you can restart the R session and run \texttt{load\_all()}. 
\item Let's try it together\dots \;NB: delete the function definition and any superfluous library call from the script used for interactive testing. 
\end{itemize}
\end{frame}

\begin{frame}[fragile]{R CMD check}
\vspace{-0.5cm}
\begin{itemize}
\itemsep 3ex
\item Every time that you modify your R package in a non-negligible way (e.g. you add a new function), you should test that all its moving parts are still working. 
\item The \texttt{R CMD check} command is the gold standard for checking R packages. 
\item We can run it from the Build panel or via the \texttt{check()} function (which is also defined in \texttt{devtools}). 
\item Let's try!
\end{itemize}
\end{frame}

\begin{frame}[fragile]{Documentation}
\vspace{-0.5cm}
\begin{itemize}
\itemsep 2ex 
\item Unfortunately, our new function doesn't have an help file:
\begin{lstlisting}
> devtools::load_all(".")
i Loading statsAndBooze
> ?parse_dates
No documentation for ‘parse_dates’ in specified packages and libraries: you could try ‘??parse_dates’
\end{lstlisting}
\item We can write a specially formatted comment right above the function definition to generate its help page via an R package named \texttt{roxygen2}. 
\item From Rstudio, open \texttt{parse.R}, place your cursor somewhere into the function definition and then click on Code -> Insert Roxygen Skeleton. 
\end{itemize}
\end{frame}

\begin{frame}[fragile]{Documentation (cont)}
\vspace{-0.5cm}
\begin{itemize}
\itemsep 2ex
\item You should see something like
\begin{lstlisting}
|\textcolor{Rcomment}{\#' Title}|
|\textcolor{Rcomment}{\#'}|
|\textcolor{Rcomment}{\#'}| |\textcolor{blue}{@param}| x 
|\textcolor{Rcomment}{\#'}|
|\textcolor{Rcomment}{\#'}| |\textcolor{blue}{@return}|
|\textcolor{Rcomment}{\#'}| |\textcolor{blue}{@export}|
|\textcolor{Rcomment}{\#'}|
|\textcolor{Rcomment}{\#'}| |\textcolor{blue}{@examples}|
parse_dates <- function(x) {
  lapply(x, lubridate::as_date)
}
\end{lstlisting}
\item Now we are going to fill all the relevant parts. 
\end{itemize}
\end{frame}

\begin{frame}[fragile]{Documentation (cont)}
\vspace{-0.5cm}
\begin{itemize}
\itemsep 2ex
\item At the end, the output should look like
\begin{lstlisting}
|\textcolor{Rcomment}{\#' Parse a list of strings into dates}|
|\textcolor{Rcomment}{\#'}|
|\textcolor{Rcomment}{\#'}| |\textcolor{blue}{@details}| |\textcolor{Rcomment}{Please notice that each date must be specified using the YYYY-MM-DD format.}|
|\textcolor{Rcomment}{\#'}|
|\textcolor{Rcomment}{\#'}| |\textcolor{blue}{@param}| dates |\textcolor{Rcomment}{A list of strings specifying dates.}|
|\textcolor{Rcomment}{\#'}|
|\textcolor{Rcomment}{\#'}| |\textcolor{blue}{@return}| |\textcolor{Rcomment}{A list with the same length as the input. The strings are converted into objects of class Date.}|
|\textcolor{Rcomment}{\#'}| |\textcolor{blue}{@export}|
|\textcolor{Rcomment}{\#'}|
|\textcolor{Rcomment}{\#'}| |\textcolor{blue}{@examples}|
|\textcolor{Rcomment}{\#' 1 + 1}|
parse_dates <- function(dates) {
  lapply(dates, lubridate::as_date)
}
\end{lstlisting}
\end{itemize}
\end{frame}

\begin{frame}[fragile]{Documentation (cont)}
\vspace{-0.5cm}
\begin{itemize}
\itemsep 3ex
\item We can run \texttt{document()} to let \texttt{roxygen2} do its magic. 
\item If you explore the NAMESPACE you should now see 
\begin{lstlisting}
|\textcolor{Rcomment}{\# Generated by roxygen2: do not edit by hand}|
export(parse_dates)
\end{lstlisting}
\item Let's run the R CMD check again. 
\end{itemize}
\end{frame}

\begin{frame}[fragile]{A minimal R package}
\vspace{-0.5cm}
\begin{itemize}
\itemsep 2ex
\item Now we have a minimal working package! We can install it by running \texttt{devtools::install()} or using the Build panel. 
\item After installing our package, we can run the following in a fresh R session
\begin{lstlisting}
|\textcolor{blue}{library}|(statsAndBooze)
beer_dates <- list(
  |\textcolor{Rcomment}{\# We can see that our function works with >= 2 people and >= 2 dates}|
  andrea = c(|\textcolor{Rstring}{"2023-03-29"}|, |\textcolor{Rstring}{"2023-03-30"}|), 
  federico = |\textcolor{Rstring}{"2023-03-30"}|,
  chiara = |\textcolor{Rstring}{"2023-03-30"}|
)
parse_dates(beer_dates)
\end{lstlisting}
\item If you don't see any error, that means our super simple package works \emoji{tada}
\end{itemize}
\end{frame}

\begin{frame}[fragile]{To infinity and beyond \emoji{rocket}}
\vspace{-0.5cm}
\begin{itemize}
\itemsep 3ex
\item Now it's time to expand our package! In fact, we said that our objective is to decide a common day for an happy hour and, currently, we are not doing that\dots 
\item In fact, for the moment we are just parsing the input constraints into a list of \texttt{Date} objects, but we are missing the key step: the organization of the happy hour!  
\item As for the previous case, it's really convenient to start running the first tests into an interactive session. 
\end{itemize}
\end{frame}

\begin{frame}[fragile]{\texttt{decide\_happy\_hour()} function}
\vspace{-0.5cm}
\begin{itemize}
\item How would you programmatically determine the common day in the following list?
\begin{lstlisting}
|\textcolor{blue}{library}|(statsAndBooze)
list_dates <- list(
  andrea = c(|\textcolor{Rstring}{"2023-03-29"}|, |\textcolor{Rstring}{"2023-03-30"}|), 
  federico = |\textcolor{Rstring}{"2023-03-30"}|,
  chiara = |\textcolor{Rstring}{"2023-03-30"}|
)
parsed_dates <- parse_dates(list_dates)
decide_happy_hour <- |\textcolor{blue}{function}|(x) {
  ... 
}
\end{lstlisting}
\item We need to recursively apply the same function to check which availabilities are shared among different people\dots\; See the next slide for a possible solution :)
\end{itemize}
\end{frame}

\begin{frame}[fragile]{\texttt{decide\_happy\_hour()} function}
\vspace{-0.5cm}
\begin{itemize}
\itemsep 2ex
\item My suggestion would be something like
\begin{lstlisting}
decide_happy_hour <- |\textcolor{blue}{function}|(x) {
  lubridate::as_date(Reduce(lubridate::intersect, x))
}
\end{lstlisting}
\item First, let's see if it works in an interactive session. 
\item If you don't see any problem, create an ad-hoc \texttt{.R} file (e.g. \texttt{decide.R}), add the new function, and document it. 
\item \textbf{Exercise for home:} Try to understand why we do need the extra call to \texttt{lubridate::as\_date}.
\end{itemize}
\end{frame}

\begin{frame}[fragile]{\texttt{decide\_happy\_hour()} function}
\vspace{-0.5cm}
\begin{itemize}
\itemsep 2ex
\item Now, after completing all the previous steps we can load\footnote{\textbf{NB:} load $\ne$ install. See \texttt{?devtools::load\_all} for more details.} our package and, in a fresh R session, retest that everything works properly: 
\begin{lstlisting}
> devtools::load_all(|\textcolor{Rstring}{"."}|)
i Loading statsAndBooze
> list_dates <- list(
  andrea = |\textcolor{Rstring}{"2023-03-30"}|, 
  federico = |\textcolor{Rstring}{"2023-03-30"}|
)
> parsed_dates <- parse_dates(list_dates)
> decide_happy_hour(parsed_dates)
[1] "2023-03-30"
\end{lstlisting}
\item If everything looks right, rerun R CMD check. Please notice that R CMD check re-documents our package. 
\end{itemize}
\end{frame}

\begin{frame}[fragile]{\texttt{decide\_happy\_hour()} function}
\vspace{-0.5cm}
\begin{itemize}
\itemsep 2ex
\item \textbf{Question:} What is the expected output of the following code? Please notice that there is no common date.  
\begin{lstlisting}
> devtools::load_all(|\textcolor{Rstring}{"."}|)
i Loading statsAndBooze
> list_dates <- list(
  andrea = |\textcolor{Rstring}{"2023-03-29"}|, 
  federico = |\textcolor{Rstring}{"2023-03-30"}|
)
> parsed_dates <- parse_dates(list_dates)
> decide_happy_hour(parsed_dates)
\end{lstlisting}
Try to formulate an hypothesis and test it running the code. 
\end{itemize}
\end{frame}

\begin{frame}[fragile]{Reinstall and more docs}
\vspace{-0.5cm}
\begin{itemize}
\itemsep 2ex
\item Now it is a good time to reinstall our R package and check again that everything works as expected. 
\item You should see following
\begin{lstlisting}
|\textcolor{blue}{library}|(statsAndBooze)
list_dates <- list(
  andrea = c(|\textcolor{Rstring}{"2023-04-01"}|, |\textcolor{Rstring}{"2023-04-02"}|), 
  federico = c(|\textcolor{Rstring}{"2023-04-02"}|, |\textcolor{Rstring}{"2023-04-03"}|), 
  chiara = |\textcolor{Rstring}{"2023-04-02"}|
)
parsed_dates <- parse_dates(list_dates)
decide_happy_hour(parsed_dates)
[1] 2023-04-02
\end{lstlisting}
\item We can also finish the docs by filling in some examples and complete the DESCRIPTION. Then CHECK!
\end{itemize}
\end{frame}

\begin{frame}[fragile]{Unit testing}
\vspace{-0.5cm}
\begin{itemize}
\itemsep 2ex
\item The example reported in the previous slide informally shows that our R package works in a particular case. 
\item Now we want to formalise our expectations into \textbf{unit tests}!
\item Why do we need unit testing? Two main reasons: 
\begin{enumerate}
\itemsep 1ex
\item We might want to test what happens with wrong inputs or other edge cases that can remain hidden for the end users;
\item We want to ensure that all aspects of our package keep working even after refactoring its main functionalities. 
\end{enumerate}
\end{itemize}
\end{frame}

\begin{frame}[fragile]{Unit testing (cont)}
\vspace{-0.5cm}
\begin{itemize}
\itemsep 1ex
\item After loading \texttt{devtools}, you can run \texttt{use\_testthat()} to setup the unit testing environment. 
\begin{lstlisting}
> use_testthat()
|\textcolor{gray}{v Setting active project to 'D:/git/statsAndBooze'}|
|\textcolor{gray}{v Adding 'testthat' to Suggests field in DESCRIPTION}|
|\textcolor{gray}{v Setting Config/testthat/edition field in DESCRIPTION to '3'}|
|\textcolor{gray}{v Creating 'tests/testthat/'}|
|\textcolor{gray}{v Writing 'tests/testthat.R'}|
|\textcolor{gray}{* Call `use\_test()` to initialize a basic test file and open it for editing.}|
\end{lstlisting}
\item Then, we can run \texttt{use\_test(\textcolor{Rstring}{<file>})} to create a new test file. So, for example, I can run \texttt{use\_test(\textcolor{Rstring}{"parse"})}. 
\begin{lstlisting}
> use_test("parse")
|\textcolor{gray}{v Writing 'tests/testthat/test-parse.R'}|
|\textcolor{gray}{* Modify 'tests/testthat/test-parse.R'}|
\end{lstlisting}
\end{itemize}
\end{frame}

\begin{frame}[fragile]{Unit testing (cont)}
\vspace{-0.5cm}
\begin{itemize}
\itemsep 2ex
\item Now we need to edit the newly created file and write our unit test(s). First, we need to provide a short sentence that summarises the objective of the test. For example
\begin{lstlisting}
test_that(|\textcolor{Rstring}{"parse\_dates(): basic functionalities work"}|, {
  expect_equal(2 * 2, 4)
})
\end{lstlisting}
\item The tests are actually run using the R package \texttt{testthat} which exports several helper functions (\texttt{expect\_length()}, \texttt{expect\_message()}, \texttt{expect\_error()}, \dots) to test different aspects of our package (equality, differences, \dots).
\item Then, we need to write the corpus of the test comparing observed output and our expectation. 
\end{itemize}
\end{frame}

\begin{frame}[fragile]{Unit testing (cont)}
\vspace{-0.5cm}
\begin{itemize}
\itemsep 2ex
\item For example: 
\begin{lstlisting}
test_that(|\textcolor{Rstring}{"parse\_dates(): basic functionalities work"}|, {
  input_strings <- list(
    andrea = |\textcolor{Rstring}{"2023-04-03"}|,
    marco = |\textcolor{Rstring}{"2023-04-03"}|
  )
  expected_dates <- list(
    andrea = lubridate::as_date(|\textcolor{Rstring}{"2023-04-03"}|),
    marco = lubridate::as_date(|\textcolor{Rstring}{"2023-04-03"}|)
  )
  expect_equal(parse_dates(input_strings), expected_dates)
})
\end{lstlisting}
\item After loading the package (\texttt{load\_all()}), we can run the new test interactively as any other R function. 
\end{itemize}
\end{frame}

\begin{frame}[fragile]{Unit testing (cont)}
\vspace{-0.5cm}
\begin{itemize}
\itemsep 2ex
\item The same procedure can be repeated for the other function
\begin{lstlisting}
> use_test(|\textcolor{Rstring}{"decide"}|)
|\textcolor{gray}{v Setting active project to 'D:/git/statsAndBooze'}|
|\textcolor{gray}{v Writing 'tests/testthat/test-decide.R'}|
|\textcolor{gray}{* Modify 'tests/testthat/test-decide.R'}|
\end{lstlisting}
\item The actual test might look like
\begin{lstlisting}
test_that(|\textcolor{Rstring}{"decide\_happy\_hour(): basic functionalities work"}|, {
  beer_dates <- list(
    andrea = lubridate::as_date(|\textcolor{Rstring}{"2023-04-03"}|),
	federico = lubridate::as_date(|\textcolor{Rstring}{"2023-04-03"}|),
	chiara = lubridate::as_date(|\textcolor{Rstring}{"2023-04-03"}|)
  )
  expect_equal(
    decide_happy_hour(beer_dates), 
	lubridate::as_date(|\textcolor{Rstring}{"2023-04-03"}|)
  )
})
\end{lstlisting}
\end{itemize}
\end{frame}

\begin{frame}[fragile]{Unit testing (cont)}
\vspace{-0.5cm}
\begin{itemize}
\itemsep 1ex
\item You should also test some pathological cases which might not be directly exposed to the regular end users
\begin{lstlisting}
test_that(|\textcolor{Rstring}{"decide\_happy\_hour(): empty intersection"}|, {
  beer_dates <- list(
    andrea = lubridate::as_date(|\textcolor{Rstring}{"2023-04-03"}|),
	marco = lubridate::as_date(|\textcolor{Rstring}{"2023-04-04"}|)
  )
  expect_equal(
    decide_happy_hour(beer_dates),
	lubridate::as_date(numeric(|\textcolor{blue}{0}|))
  )
})
\end{lstlisting}
\item Similarly, you could develop a test to control the behaviour of the function in case of misspecified inputs. 
\item For example: what is the expected output when one or more of the input Date(s) is NA?
\end{itemize}
\end{frame}

\lstset{escapechar={£}}

\begin{frame}[fragile]{Unit testing (cont)}
\vspace{-0.5cm}
\begin{itemize}
\itemsep 2ex
\item The function \texttt{test()} (which is also defined in \texttt{devtools}) automatically runs all tests in a package and returns an informative output (with a funny comment on the results)
\begin{lstlisting}
> test()
i Testing statsAndBooze
v | F W S  OK | Context
v |         2 | decide
v |         1 | parse
	
== Results ===============================
Duration: 0.5 s
		
[ FAIL 0 | WARN 0 | SKIP 0 | £\textcolor{Rcomment}{PASS}£ 3 ] 
		
£\textcolor{Rcomment}{You are a coding rockstar!}£
\end{lstlisting}
\item The same behaviour occurs when you run R CMD check.
\end{itemize}
\end{frame}

\begin{frame}
\vspace{2cm}
\begin{center}
\Huge
\textbf{THE END!}
\end{center}
\vspace{1.5cm}
\epigraph{E quindi uscimmo a riveder le stelle}{Dante Alighieri, Inferno XXXIV, 139}
\end{frame}
\end{document}
